One-stage point transect distance sampling using iterated integrated nested Laplace approximations

### 2019-12-10 paper-v2.tex
actually this one to Janine

\begin{abstract}
Distance sampling methods aim to estimate the size of animal populations.  Here we present an analysis of point transect distance sampling data collected on a tropical forest bird, presenting an entire workflow from model conception, inference procedure, model evaluation to communication of results, at each stage discussing the challenges involved and our strategies to address them.

New developments inlcude a new inference method for one-stage distance sampling models within a Bayesian framework based on iterated fits of latent Gaussian models fitted using integrated nested Laplace approximations \citep{rue_approximate_2009} to simultaneously estimate the detectability and spatial distribution of populations of wild animals.  We take a point process perspective, modelling the data as a thinned log-Gaussian Cox process using computationally efficient Gaussian Markov random field approximations to Gaussian random fields implemented via a stochastic partial differential equation approach \citep{lindgren_explicit_2011}.  The Bayesian setting allows sampling from the joint posterior of all model parameters and so the resulting model outputs naturally incorporate uncertainty in the observation model estimates.  We discuss potential issues with common ways to represent uncertainty in mapped estimates of species distributions and advocate for the \textit{a priori} selection of relevant thresholds and uncertainty levels.  We apply the methods of excursion sets and excursion functions \citep{bolin_excursion_2015} as an example of this approach.
\end{abstract}

@@

[[Need to update this and the figure with the extended data area]]

Hakalau was established in 1985 to conserve 15,390-ha of montane forest habitat for native forest birds and rainforest plants. Annual forest bird surveys were initiated in 1987 to determine population status and track trends in abundance. Survey points were established along 15 transects following a systematic, random design with points approximately 150 m apart on transects located either 500 or 1,000 m apart. Following \cite{camp_population_2010, camp_statespace_2016}, we restrict our study area to the 3,061 ha open-forest stratum of Hakalau.  The open-forest stratum was previously heavily grazed, and since the removal of cattle in 1988 regeneration has proceeded naturally \citep{maxfield_hakalau_1998}. To the north the study area follows the refuge boundary while to the east it is bounded by a fence line (Fig. \ref{fig:2002studyareapointspt}). The southern boundary was modified from \cite{camp_population_2010} to exclude the non-sampled forested portion of their study area. The west side of the study area is bounded by pasture that is dominated by grass and is unsuitable habitat for \akepa.

>>>>

\begin{abstract}
Distance sampling methods aim to estimate the size and spatial distribution of populations of wild animals.  Here we present an analysis of point transect distance sampling data collected on a tropical forest bird.  We consider an entire workflow, from model conception, inference procedure, model evaluation to communication of results, at each stage discussing the challenges involved and our strategies to address them.

New developments inlcude an inference method for one-stage distance sampling models within a Bayesian framework based on iterated fits of latent Gaussian models fitted using integrated nested Laplace approximations \citep{rue_approximate_2009}.  This approach simultaneously estimates the detectability and spatial distribution of populations of wild animals within a joint model.  We take a point process perspective, modelling the data as a thinned log-Gaussian Cox process using computationally efficient Gaussian Markov random field approximations to Gaussian random fields implemented via a stochastic partial differential equation approach \citep{lindgren_explicit_2011}.  The Bayesian setting allows sampling from the joint posterior of all model parameters and so the resulting model outputs naturally incorporate uncertainty in the observation model estimates.  We discuss potential issues with common ways to represent uncertainty in mapped estimates of species distributions and advocate for the \textit{a priori} selection of relevant thresholds and uncertainty levels.  We apply the methods of excursion sets and excursion functions \citep{bolin_excursion_2015} as an example of this approach.
\end{abstract}

@@

Hakalau was established in 1985 to conserve 15,390-ha of montane forest habitat for native forest birds and rainforest plants. Annual forest bird surveys were initiated in 1987 to determine population status and track trends in abundance. Survey points were established along 15 transects following a systematic, random design with points approximately 150 m apart on transects located either 500 or 1,000 m apart. We restrict our study area to the open-forest and closed-forest strata of Hakalau, an extension of the area considered in \cite{camp_population_2010, camp_statespace_2016} which only considered did not include the closed-forest strata which was not sampled in the early years of the \akepa{} surveys.  The open-forest stratum was previously heavily grazed, and since the removal of cattle in 1988 regeneration has proceeded naturally \citep{maxfield_hakalau_1998}. To the north the study area follows the refuge boundary while to the east it is bounded by a fence line (Fig. \ref{fig:2002studyareapointspt}). The southern boundary was modified from \cite{camp_population_2010} to exclude the non-sampled forested portion of their study area. The west side of the study area is bounded by pasture that is dominated by grass and is unsuitable habitat for \akepa.
